Density-dependent selection and the limits of relative fitness

### 2018-08-31 manuscript.tex
l

Selection is commonly described by assigning constant relative fitness values to genotypes. Yet population density is often regulated by crowding. Relative fitness may then depend on density, and selection can change density when it acts on a density-regulating trait. When strong density-dependent selection acts on a density-regulating trait, selection is no longer describable by density-independent relative fitnesses, even in demographically stable populations. These conditions are met in most previous models of density-dependent selection (e.g.~``$K$-selection'' in the logistic and Lotka-Volterra models), suggesting that density-independent relative fitnesses must be replaced with more ecologically explicit absolute fitnesses unless selection is weak. Here we show that density-independent relative fitnesses can also accurately describe strong density-dependent selection under some conditions. We develop a novel model of density-regulated population growth with three ecologically intuitive traits: fecundity, mortality, and competitive ability. Our model, unlike the logistic or Lotka-Volterra, incorporates a density-dependent juvenile ``reproductive excess'', which largely decouples density-dependent selection from the regulation of density. We find that density-independent relative fitnesses accurately describe strong selection acting on any one trait, even fecundity, which is both density-regulating and subject to density-dependent selection. Our findings suggest that deviations from demographic equilibrium pose the most serious difficulties for relative fitness models. In such cases our model offers a possible alternative to relative fitness.

\noindent (210 words)

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{strengthofselection.pdf}
\caption{\label{fig:strengthofselection} (a) Change in the selection coefficient between the beginning and end of a sweep of a type that experiences proportionally $1-\epsilon$ fold fewer crowding-induced deaths. The population is in demographic equilibrium at the start and end of the sweep.}
\end{figure}

@@

Summarizing the properties of selection in our variable-density lottery model: (i) $c$-selection is density-dependent, but $c$ does not regulate density; (ii) $d$ regulates density, but $d$-selection is density-independent; (iii) $b$ regulates density and $b$-selection is density-dependent. Yet, despite the differences between $b$, $c$ and $d$,  selection in a constant environment that only involves one of these traits obeys the density-independent relative fitness description of selection almost exactly (that is, $(W_2-W_1)/\overline{W}$ in Eq.~\eqref{eq:canonicaldisc} is effectively independent of density). This density-independence breaks down when strong selection acts on more than one of $b$, $c$ and $d$ (Fig.~\ref{fig:bdsweep}). The $c$ and $d$ traits exemplify the two distinct directions in which density and selection can interact: selection may depend on density, and density may change in response to ongoing selection \citep{prout_1980}. The combination of both is necessary to invalidate the constant-$s$ approximation. Remarkably, the $b$ trait demonstrates that the combination is not sufficient; the density-dependence of $b$-selection effectively disappears over equilibrium-to-equilibrium $b$-sweeps.

Selection in the variable-density lottery is quite different from classical density-dependent selection (see ``Introduction'' and ``The response of density to selection; $c$-selection versus $K$-selection''). In the latter, only one life-history stage is represented, and the effects of crowding appear as a reduction in absolute fitness that only depends on the type densities at this life-history stage (e.g. the $n_i^2$ and $n_in_j$ terms in the Lotka-Volterra equation). Selection in crowded populations takes broadly one of two forms: selection for greater carrying capacity ($K$-selection) or selection on competition coefficients ($\alpha$-selection). These are both ``$\delta$-like''  in the sense that selection depends on density and also causes density to change ($\delta$ is defined in Eq.~\eqref{eq:simplebirthdeath}). Strong selection is therefore sufficient for Eq.~\eqref{eq:canonicalcont} to break down (Fig.~\ref{fig:strengthofselection}), and no distinction is made between density-regulating and density-dependent traits.

@@

Looking beyond the variable-density lottery, it is not clear which forms of crowding-induced selection are more likely to occur in nature. Even if reproductive excesses are ubiquitous, strictly relative $c$-like traits could pleiotropically interact with density-regulating traits so often that $\delta$-like behavior is prevalent. For instance, in the case in the case of well-mixed indirect exploitation competition for consumable resources, the $R^*$ rule suggests that competitive ability is intimately linked to equilibrium resource density, and hence that $\delta$-like behavior would be prevalent. However, this conclusion is sensititive to the assumptions of well-mixed resource competition models. Spatial localization of consumable resources (e.g. for plants due to restricted movement of nutrients through soils) will tend to create territorial contests similar to the lottery model, where resource competition only occurs locally and can be sensitive to contingencies such as the timing of propagule arrival \citep{bolker_99}. In this case, resource competition is effectively subsumed into a territorial competitive ability trait akin to $c$, which would likely affect $N$ much more weakly than suggested by the $R^*$ rule (assuming no pleiotropic interactions with $b$ or $d$).

Moreover, even in well-mixed populations, competition does not only involve indirect exploitation of shared resources, but also direct interference. Interference competition can dramatically alter the dynamics of resource exploitation \citep{case_1974,amarasekare_2002}, and is more likely than the exploitation of shared resource pools to involve relative contests akin to $c$-selection. For instance, sexual selection can be viewed as a form of relative interference competition between genotypes. Thus, \textit{a priori} we should not expect crowding in nature to only involve selection that is $\delta$-like. Other forms of selection like $c$-selection (that is, strictly relative traits in density-regulated populations) are also likely to be important. Note that in the classical density-dependent selection literature, interference competition is closely associated with $\alpha$-selection and the idea that selection need not affect equilibrium density \citep{gill_1974}. However, $\alpha$-selection does transiently affect population density and therefore retains $\delta$-like features.

The above findings suggest that the largest deviations from the approximation of density-independent selection (as represented by Eqs.~\eqref{eq:canonicalcont} and \eqref{eq:canonicaldisc}) will occur in populations far from demographic equilibrium e.g.~as a result of a temporally-variable environment. While transient deviations from demographic equilibrium driven by the appearance of new types can also cause the density-independent approximation to break down, this requires strong selection that is both density-dependent and affects a density-regulating trait (and, as exemplified by $b$-selection, even then the approximation may hold). By contrast, temporally-variable environments can dramatically alter frequency trajectories for individual sweeps (e.g. Fig.~9.5 in \cite{otto_2011}; Fig.~5 in \cite{mallet_2012}), as well as the long-term outcomes of selection \citep{lande_2009}.

>>>>

Selection is commonly described by assigning constant relative fitness values to genotypes. Yet population density is often regulated by crowding. Relative fitness may then depend on density, and selection can change density when it acts on a density-regulating trait. When strong density-dependent selection acts on a density-regulating trait, selection is no longer describable by density-independent relative fitnesses, even in demographically stable populations. These conditions are met in most previous models of density-dependent selection (e.g.~``$K$-selection'' in the logistic and Lotka-Volterra models), suggesting that density-independent relative fitnesses must be replaced with more ecologically explicit absolute fitnesses unless selection is weak. Here we show that density-independent relative fitnesses can also accurately describe strong density-dependent selection under some conditions. We develop a novel model of density-regulated population growth with three ecologically intuitive traits: fecundity, mortality, and competitive ability. Our model, unlike the logistic or Lotka-Volterra, incorporates a density-dependent juvenile ``reproductive excess'', which largely decouples density-dependent selection from the regulation of density. We find that density-independent relative fitnesses accurately describe strong selection acting on any one trait, even fecundity, which is both density-regulating and subject to density-dependent selection. Pleiotropic interactions between these traits recovers the familiar $K$-selection behavior. In such cases, or when the population is maintained far from demographic equilibrium, our model offers a possible alternative to relative fitness.

\noindent (215 words)

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{strengthofselection.pdf}
\caption{\label{fig:strengthofselection} Change in the selection coefficient between the beginning and end of a sweep of a type that experiences proportionally $1-\epsilon$ fold fewer crowding-induced deaths. The population is in demographic equilibrium at the start and end of the sweep.}
\end{figure}

@@

Summarizing the properties of selection in our variable-density lottery model: (i) $c$-selection is density-dependent, but $c$ does not regulate density; (ii) $d$ regulates density, but $d$-selection is density-independent; (iii) $b$ regulates density and $b$-selection is density-dependent. Yet, despite the differences between $b$, $c$ and $d$,  selection in a constant environment that only involves one of these traits obeys the density-independent relative fitness description of selection almost exactly (that is, $(W_2-W_1)/\overline{W}$ in Eq.~\eqref{eq:canonicaldisc} is approximately independent of density). This density-independence breaks down when strong selection acts on more than one of $b$, $c$ and $d$ (Fig.~\ref{fig:bdsweep}). The $c$ and $d$ traits exemplify the two distinct directions in which density and selection can interact: selection may depend on density, and density may change in response to ongoing selection \citep{prout_1980}. The combination of both is necessary to invalidate the constant-$s$ approximation. Remarkably, the $b$ trait demonstrates that the combination is not sufficient; the density-dependence of $b$-selection effectively disappears over equilibrium-to-equilibrium $b$-sweeps.

@@

In practice, traits like $b$, $c$ and $d$ will often have pleiotropic interactions which mean that adaptive sweeps will behave similarly to the familiar ``$\delta$ sweeps'' (``Density regulating traits under strong selection''). Thus, our analysis of the variable density lottery does not suggest that Eq.~\eqref{eq:canonicalcont} and \eqref{eq:canonicaldisc} should apply more broadly than previously thought. Rather, $b$, $c$ and $d$ represent a possible set of idealized fitness components mediating the interplay between selection and density in density-regulated populations. The conceptual distinctions they highlight will hopefully be useful regardless of the biological realism of the variable density lottery model.

Apart from familiar examples such as reef fishes \citep{chesson_1981}, many biological systems do not obviously satisfy the assumptions of our variable density lottery. However, even if competition occurs primarily via consumable resource exploitation, spatial localization of consumable resources (e.g. for plants due to restricted movement of nutrients through soils) will tend to create territorial contests similar to the lottery model, where resource competition only occurs locally and can be sensitive to contingencies such as the timing of propagule arrival \citep{bolker_99}. In this case, resource competition is effectively subsumed into a territorial competitive ability trait akin to $c$, which would likely affect $N$ much more weakly than suggested by the $R^*$ rule (assuming no pleiotropic interactions with $b$ or $d$). Moreover, even in well-mixed populations, competition does not only involve indirect exploitation of shared resources, but also direct interference. Interference competition can dramatically alter the dynamics of resource exploitation \citep{case_1974,amarasekare_2002}, and is more likely than the exploitation of shared resource pools to involve relative contests akin to $c$-selection. For instance, sexual selection can be viewed as a form of relative interference competition between genotypes.

In the analysis presented here we have restricted our attention to selection in demographically stable populations. The largest deviations from the approximation of density-independent selection (as represented by Eqs.~\eqref{eq:canonicalcont} and \eqref{eq:canonicaldisc}) will likely occur in populations far from demographic equilibrium e.g.~as a result of a temporally-variable environment. This is because extremely strong selection to change population density by an amount comparable to environmental varability (see Fig.~\ref{fig:strengthofselection}). By contrast, temporally-variable environments can dramatically alter frequency trajectories for individual sweeps (e.g. Fig.~9.5 in \cite{otto_2011}; Fig.~5 in \cite{mallet_2012}), as well as the long-term outcomes of selection \citep{lande_2009}.
